---
title: "Data types"
author: "Daniel Anderson "
date: "Week 1, Class 2 "
output:
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
    includes:
      in_header: "load-feather.html"
---

```{r include = FALSE, results = "asis"}
source(here::here("static", "slides", "slide-setup.R"))
```

`r setup("w1p2")`

---
# Agenda
* Finishing up on coercion

* Attributes

* Missing values

* Intro to lists

* Subsetting


---
# Learning objectives
* Understand the fundamental difference between lists and atomic vectors


--
* Understand how atomic vectors are coerced, implicitly or explicitly


--
* Understand various ways to subset vectors, and how subsetting differs for lists


--
* Understand what an attribute is, and how to set and modify attributes

---
# Pop quiz
Without actually running the code, predict which type each of
the following will coerce to.

```{r imp-coerce2-echo, eval = FALSE}
c(TRUE, 1L, 0L, "False")
c(1L, FALSE)
c(7L, 6.23, "eight")
c(1.25, TRUE, 4L)
```

`r countdown::countdown(minutes = 1)`

---
# Answers

```{r imp-coerce2-eval}
typeof(c(TRUE, 1L, 0L, "False"))
typeof(c(1L, FALSE))
typeof(c(7L, 6.23, "eight"))
typeof(c(1.25, TRUE, 4L))
```

---
# Challenge

### Work with a partner
One of you share your screen: 
* Create four atomic vectors, one for each of the fundamental types

* Combine two or more of the vectors. Predict the implicit coercion of each.

* Apply explicit coercions, and predict the output for each.

(basically quiz each other)

`r countdown::countdown(minutes = 8)`

---
class: inverse-blue middle
# Attributes

---
# Attributes
* What are attributes?


--
  + metadata... what's metadata?


--
  + Data about the data


---
# Other data types

Atomic vectors by themselves make up only a small fraction of the total number of data types in R


--
### What are some other data types?

--
* Data frames 

--
* Matrices & arrays 

--
* Factors 

--
* Dates 


--
Remember, atomic vectors are the atoms of R. Many other data structures are
built from atomic vectors. 

* We use attributes to create other data types from atomic vectors

---
# Attributes

### Common
* Names
* Dimensions

### Less common
* Arbitrary metadata


---
# Examples
### Please follow along!
* See **all** attributes associated with a give object with `attributes`

```{r attributes-iris}
library(palmerpenguins)
attributes(penguins[1:50, ]) # limiting rows just for slides
```

---
```{r iris-head}
head(penguins)
```

---
# Get specific attribute
* Access just a single attribute by naming it within `attr`

```{r iris-class}
attr(penguins, "class")
attr(penguins, "names")
```


--
Note - this is not generally how you would pull the names attribute. Rather, you would use `names()`. 

---
# Be specific
* Note in the prior slides, I'm asking for attributes on the entire data frame. 

* Is that what I want?... maybe. But the individual vectors may have attributes as well

--
```{r attr-species}
attributes(penguins$species)
attributes(penguins$bill_length_mm)
```


---
# Set attributes
* Just redefine them within `attr`

```{r attr-redefine}
attr(penguins$species, "levels") <- c("Big one", 
                                      "Little one", 
                                      "Funny one")

head(penguins)
```

Note - you would generally not define levels this way, but it is a general method for modifying attributes.

---
# Dimensions

* Let's create a matrix (please do it with me)

```{r mat1}
m <- matrix(1:6, ncol = 2)
m
```

* Notice how the matrix fills

--
* Check out the attributes

```{r mat1-attributes}
attributes(m)
```

---
# Modify the attributes
* Let's change it to a 2 x 3 matrix, instead of 3 x 2 (you try first)

--
```{r mat1-modify-attr}
attr(m, "dim") <- c(2, 3)
m
```

--
* is this the result you expected?

---
# Alternative creation
* Create an atomic vector, assign a dimension attribute

```{r v-to-m}
v <- 1:6
v
attr(v, "dim") <- c(3, 2)
v
```

---
# Aside
* What if we wanted it to fill by row?

.pull-left[
```{r by-row1}
matrix(6:13, 
       ncol = 2, 
       byrow = TRUE) #<<
vect <- 6:13
dim(vect) <- c(2, 4)
vect
```
]

.pull-right[
```{r by-row2}
t(vect)
```
]

---
# Names
* The following (this slide and the next) are equivalent 

```{r dimnames1}
attr(v, "dimnames") <- list(c("the first", "second", "III"),
                            c("index", "value"))
v
```

---
# Names

```{r dimnames2}
v2 <- 1:6
attr(v2, "dim") <- c(3, 2)
rownames(v2) <- c("the first", "second", "III")
colnames(v2) <- c("index", "value")
v2
```

---
# Arbitrary metadata
* I don't use this often (wouldn't recommend you do either)

```{r }
attr(v, "matrix_mean") <- mean(v)
v
attr(v, "matrix_mean")
```

* Note that *anything* can be stored as an attribute (including matrices or data frames, etc.)

---
# Matrices vs Data frames

Usually we want to work with data frames because they represent our data better.

Sometimes a matrix is more efficient because you can operat on the **entire** matrix at once.

```{r }
set.seed(42)
m <- matrix(rnorm(100, 200, 10), ncol = 10)
m
```

---
```{r }
sum(m)
mean(m)
rowSums(m)
colSums(m)

# standardize the matrix
z <- (m - mean(m)) / sd(m)
```

---
```{r }
z
```

---
# Stripping attributes
* Many operations will strip attributes (generally why it's not a good idea to store important things in them)

.pull-left[
```{r stripped1}
v
rowSums(v)
```
]

--
.pull-right[
```{r stripped2}
attributes(rowSums(v))
```

* Generally `names` are maintained

* Sometimes, `dim` is maintained, sometimes not 

* All else is stripped

]



---
# More on `names`

* The `names` attribute corresponds to the individual elements within a vector

```{r names1}
names(v)
names(v) <- letters[1:6]
v
```

---
* Perhaps more straightforward

```{r names2}
v3a <- c(a = 5, b = 7, c = 12)
v3a
names(v3a)
attributes(v3a)
```

---
# Alternatives

```{r names3}
v3b <- c(5, 7, 12)
names(v3b) <- c("a", "b", "c")
v3b

v3c <- setNames(c(5, 7, 12), c("a", "b", "c"))
v3c
```


--
* Note that `names` is **not** the same thing as `colnames`, but, somewhat confusingly, both work to rename the variables (columns) of a data frame. We'll talk more about why this is momentarily.


---
# Why names might be helpful

```{r}
v
```

.pull-left[
```{r}
v["b"]
```
]

.pull-right[
```{r }
v["e"]
```
]

---
# Implementation of factors
### Quickly

```{r factor}
fct <- factor(c("a", "a", "b", "c"))
typeof(fct)
attributes(fct)
str(fct)
```

---
# Implementation of dates
### Quickly

```{r dates}
date <- Sys.Date()
typeof(date)
attributes(date)
attributes(date) <- NULL
date
```

* This number represents the days passed since January 1, 1970, known as the Unix epoch.

---
# Missing values

* Missing values breed missing values

```{r missing1}
NA > 5
NA * 7
```

--

* What about this one?

```{r missing2, eval = FALSE}
NA == NA
```

--

```{r missing2-eval, echo = FALSE}
NA == NA
```


--
It is correct because there's no reason to presume that one missing
value is or is not equal to another missing value.

---
# When missing values don't propagate

```{r missing-non-propagation}
NA | TRUE
x <- c(NA, 3, NA, 5)
any(x > 4)
```

---
# How to test missingness?

* We've already seen the following doesn't work

```{r missing3}
x == NA
```

--
* Instead, use `is.na`

```{r missing4}
is.na(x)
```

* When does this regularly come into play?


---
class: inverse-blue middle
# Lists


---
# Lists
* Lists are vectors, but not *atomic* vectors

* Fundamental difference - each element can be a different type

```{r list1}
list("a", 7L, 3.25, TRUE)
```

---
# Lists

.pull-left[

* Technically, each element of the list is a vector, possibly atomic

* The prior example included all *scalars*, which are vectors of length 1.

* Lists do not require all elements to be the same length

]


.pull-right[

```{r list2}
l <- list(
  c("a", "b", "c"),
  rnorm(5),
  c(7L, 2L),
  c(TRUE, TRUE, FALSE, TRUE)
)
l
```
]


---
# Check the list

```{r list-attr}
typeof(l)
attributes(l)
str(l)
```

---
# Data frames as lists
* A data frame is just a special case of a list, where all the elements are of the same length.

.pull-left[
```{r list-df1}
l_df <- list(
  a = c("red", "blue"),
  b = rnorm(2),
  c = c(7L, 2L),
  d = c(TRUE, FALSE)
)
l_df
```
]

.pull-right[
```{r list-df2}
data.frame(l_df)
```
]


---
class: inverse-red middle
# Subsetting Lists

---
# A nested list
Lists are often complicated objects. Let's create a somewhat complicated one

```{r }
x <- c(a = 3, b = 5, c = 7)
l <- list(
  x = x,
  x2 = c(x, x),
  x3 = list(
    vect = x,
    squared = x^2,
    cubed = x^3)
)
```

---
# Subsetting lists
Multiple methods
* Most common: `$`, `[`, and `[[`

.pull-left[
```{r subset-lists-left}
l[1]
typeof(l[1])

l[[1]]
```
]


.pull-right[
```{r subset-lists-right}
typeof(l[[1]])

l[[1]]["c"]
```
]

---
# Named list
* Because the elements of the list are named, we can use `$`

```{r dollar-subset}
l$x2
l$x3
```


---
# Subsetting nested lists
* Multiple `$` if all named

```{r dollar-subset-nested}
l$x3$squared
```

* Note this doesn't work on named elements of an atomic vector, just the named elements of a list

```{r dollar-subset-error, error = TRUE}
l$x3$squared$b
```


---
But we could do something like...

```{r dollar-subset-success}
l$x3$squared["b"]
```

---
# Alternatives

* You can always use logical

* Indexing works too

.pull-left[
```{r alternative-subset-list1}
l[c(TRUE, FALSE, TRUE)]
```
]

.pull-right[
```{r alternative-subset-list2}
l[c(1, 3)]
```
]

---
# Careful with your brackets

```{r bracket-fail, error = TRUE}
l[[c(TRUE, FALSE, FALSE)]]
```

* Why doesn't the above work?

---
## Subsetting in multiple dimensions

* Generally we deal with 2d data frames

* If there are two dimensions, we separate the `[` subsetting with a comma

```{r head-mtcars}
head(mtcars)
mtcars[3, 4]
```

---
# Empty indicators
* An empty indicator implies "all"


--
### Select the entire fourth column

```{r mtcars-col4}
mtcars[ ,4]
```

--
### Select the entire 4th row

```{r mtcars-row4}
mtcars[4, ]
```

---
# Data types returned

* By default, each of the prior will return a vector, which itself can be subset

The following are equivalent

```{r mtcars-row4-mpg1}
mtcars[4, c("mpg", "hp")]
```

```{r mtcars-row4-mpg2}
mtcars[4, ][c("mpg", "hp")]
```

---
# Return a data frame
* Often, you don't want the vector returned, but rather the modified data frame. 

* Specify `drop = FALSE`

```{r drop-false}
mtcars[ ,4]
mtcars[ ,4, drop = FALSE]
```

---
# tibbles
* Note dropping the data frame attribute is the default for a `data.frame` but .b[.bolder[NOT]] a `tibble`.

```{r tibble-nodrop-att}
mtcars_tbl <- tibble::as_tibble(mtcars)
mtcars_tbl[ ,4]
```

---
# You can override this

```{r tibble-drop-att}
mtcars_tbl[ ,4, drop = TRUE]
```

---
# More than two dimensions
* Depending on your applications, you may not run into this much

```{r array}
array <- 1:12
dim(array) <- c(2, 3, 2)
array
```

---
# Subset array
### Select just the second matrix

--

```{r array-select-mat}
array[ , ,2]
```

--

### Select first column of each matrix

--

```{r array-select-col}
array[ ,1, ]
```

---
# Back to lists
### Why are they so useful?
* Fairly obviously, they're much more flexible

* Often returned by functions, for example, `lm`

```{r lm1}
m <- lm(mpg ~ hp, mtcars)
str(m)
```

---
# Summary
* Atomic vectors must all be the same type

  + implicit coercion occurs if not (and you haven't specified the coercion
  explicitly)

* Lists are also vectors, but not atomic vectors
  
  + Each element can be of a different type and length

  + Incredibly flexible, but often a little more difficult to get the hang of, particularly with subsetting


---
class: inverse-green middle
# Next time
### Loops with base R

