---
title: "List columns"
author: "Daniel Anderson "
date: "Week 4, Class 2"
output:
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
    includes:
      in_header: "load-feather.html"
---

```{r include = FALSE, results = "asis"}
source(here::here("static", "slides", "slide-setup.R"))
xaringanExtra::use_clipboard()
knitr::opts_chunk$set(fig.width = 13, 
                      message = FALSE, 
                      warning = FALSE)

library(tidyverse)
theme_set(theme_minimal(20))
update_geom_defaults("point", list(size = 3))
```

`r setup("w4p2")`

---
# Agenda
* Review Lab 2

* Introduce list columns

* Contrast: 
  + `group_by() %>% nest() %>% mutate() %>% map()` with 
  + `nest_by() %>% summarize()`

* In-class midterm (last 20 minutes)

---
# Learning objectives
* Understand list columns and how they relate to `base::split`

* Fluently nest/unnest data frames

* Understand why `tidyr::nest` can be a powerful framework (data frames) and
when `tidyr::unnest` can/should be used to move out of nested data frames and
into a standard data frame.

---
class: inverse-red center middle
# Review Lab 2

---
# Setup
## Please follow along

First import the data

```{r import-data}
library(tidyverse)
library(fs)
files <- dir_ls(here::here("data", "pfiles_sim"), 
                glob = "*.csv")
d <- files %>%
	map_df(read_csv, .id = "file")
```

---
# Parse file data
```{r }
d <- d %>%
  mutate(
    file = str_replace_all(
	  file, 
	  here::here("data", "pfiles_sim"), 
	  ""
    ),
    grade = str_replace_all(file, "/g(\\d?\\d).+", "\\1"),
    grade = as.integer(grade),
    year = str_replace_all(
      file, 
      ".+files(\\d\\d)_sim.+", 
      "\\1"
    ),
    year = as.integer(year),
    content = str_replace_all(
      file, 
      "/g\\d?\\d(.+)pfiles.+", 
      "\\1"
    )
)
```

---
# Select variables

```{r }
d <- d %>% 
	select(ssid, grade, year, content, testeventid, 
	       asmtprmrydsbltycd, asmtscndrydsbltycd, Entry:WMLE)
```

---
# Comparing models
Let's say we wanted to fit/compare a set of models for each content area

1. `lm(Theta ~ asmtprmrydsbltycd)`

1. `lm(Theta ~ asmtprmrydsbltycd + asmtscndrydsbltycd)`

1. `lm(Theta ~ asmtprmrydsbltycd * asmtscndrydsbltycd)`

---
# Data pre-processing
* The disability variables are stored as numbers, we need them as factors
* We'll make the names easier in the process

```{r disab-recode}
d <- d %>%
	mutate(primary = as.factor(asmtprmrydsbltycd),
	       secondary = as.factor(asmtscndrydsbltycd))
```

If you're interested in what the specific codes refer to, see [here](https://www.newberg.k12.or.us/district/eligibility-codes-and-requirements).

---
# Split the data

The base method we've been using...

```{r split}
splt_content <- split(d, d$content)
str(splt_content)
```

---
# We could use this method

```{r fit-models-split}
m1 <- map(
  splt_content, 
  ~lm(Theta ~ asmtprmrydsbltycd, data = .x)
)

m2 <- map(
  splt_content, 
  ~lm(Theta ~ asmtprmrydsbltycd + asmtscndrydsbltycd, 
      data = .x)
)

m3 <- map(
  splt_content, 
  ~lm(Theta ~ asmtprmrydsbltycd * asmtscndrydsbltycd, 
      data = .x)
)
```

* Then conduct tests to see which model fit better, etc.


---
# Alternative
* Create a data frame with a list column

```{r list-column}
by_content <- d %>%
  group_by(content) %>% 
  nest()
by_content
```

---
# What's going on here?

```{r list-column-listed}
str(by_content$data)
```

---
# Explore a bit

```{r map-data1}
map_dbl(by_content$data, nrow)
map_dbl(by_content$data, ncol)
map_dbl(by_content$data, ~mean(.x$Theta))
```

---
# It's a data frame!

We can add these summaries if we want

```{r n-by-content}
by_content %>%
	mutate(n = map_dbl(data, nrow))
```

---
# `map_*`
* Note on the previous example we used `map_dbl` and we got a vector in return. 

* What would happen if we just used `map`?

--

```{r n-by-content-list}
by_content %>%
	mutate(n = map(data, nrow))
```

---
# Let's fit a model!

```{r list-column-model}
by_content %>%
	mutate(m1 = map(data, ~lm(Theta ~ primary, data = .x)))
```

---
# Extract the coefficients

```{r }
by_content %>%
  mutate(
    m1 = map(data, ~lm(Theta ~ primary, data = .x)),
    coefs = map(m1, coef)
)
```

---
# Challenge

* Continue with the above, but output a data frame with three columns:
`content`, `intercept`, and `TBI` (which is code 74). 

* In other words, output the
mean score for students who were coded as not having a disability (code 0),
along with students coded as having TBI.

`r countdown::countdown(4)`

---

```{r coefs}
by_content %>%
  mutate(
    m1 = map(data, ~lm(Theta ~ primary, data = .x)),
    coefs = map(m1, coef),
    no_disab = map_dbl(coefs, 1),
    tbi = no_disab + map_dbl(coefs, "primary74")
  ) %>%
  select(content, no_disab, tbi)
```

--
Note - I wouldn't have neccesarily expected you to add `no_disab` to the TBI coefficient.

---
# Compare models
* Back to our original task - fit all three models

### You try first

1. `lm(Theta ~ primary)`

1. `lm(Theta ~ primary + secondary)`

1. `lm(Theta ~ primary + secondary + primary:secondary)`

`r countdown::countdown(4)`

---
# Model fits

```{r three-mods}
mods <- by_content %>%
  mutate(
    m1 = map(data, ~lm(Theta ~ primary, data = .x)),
    m2 = map(data, ~lm(Theta ~ primary + secondary, data = .x)),
    m3 = map(data, ~lm(Theta ~ primary * secondary, data = .x))
)
mods
```

---
# Brief foray into parallel iterations
The `stats::anova` function can compare the fit of two models

--
### Pop Quiz

How would we extract just ELA model 1 and 2?

--

.pull-left[
```{r ela-mods1}
mods$m1[[1]]
```
]

.pull-right[
```{r ela-mods2}
mods$m2[[1]]
```
]

---
# Which fits better?

```{r anova}
compare <- anova(mods$m1[[1]], mods$m2[[1]])
compare
```

---
# `map2`

* Works the same as `map` but iterates over two vectors concurrently

* Let's compare model 1 and 2

--
```{r mod_compare1}
mods %>%
  mutate(comp12 = map2(m1, m2, anova))
```

--
Perhaps not terrifically helpful


---
# Back to our `anova` object
* Can we pull out  useful things?

```{r str-comp}
str(compare)
```

--
Try pulling out the $p$ value

---
# Extract $p$ value
* *Note - I'd recommend looking at more than just a p-value, but I do think this
is useful for a quick glance*

```{r extract-p1}
compare$`Pr(>F)`
compare[["Pr(>F)"]]
```

--

```{r extract-p2}
compare$`Pr(>F)`[2]
compare[["Pr(>F)"]][2]
```

---
# All p-values

*Note - this is probably the most compact syntax, but that doesn't mean it's the
most clear*

```{r pull-pvals12}
mods %>%
  mutate(comp12 = map2(m1, m2, anova),
         p12 = map_dbl(comp12, list("Pr(>F)", 2)))
```

---
# Slight alternative
* Write a function that pulls the p-value from model comparison objects

```{r fun-compare}
extract_p <- function(anova_ob) {
  anova_ob[["Pr(>F)"]][2]
}
```


--
* Loop this function through the anova objects

---
```{r pull-pvals12-fun}
mods %>%
  mutate(comp12 = map2(m1, m2, anova),
         p12 = map_dbl(comp12, extract_p))
```

---
class: inverse-red middle
# An alternative
## Conducting operations by row

---
# Operations by row
The `dplyr::rowwise()` function fundamentally changes the way a `tibble()` behaves

```{r }
df <- tibble(name = c("Me", "You"), x = 1:2, y = 3:4, z = 5:6)
```

.pull-left[
```{r }
df %>% 
  mutate(m = mean(c(x, y, z)))
```
]

.pull-right[
```{r }
df %>% 
  rowwise() %>% 
  mutate(m = mean(c(x, y, z)))
```
]

---
# Add a group & summarize

```{r }
df %>% 
  rowwise(name) %>% 
  summarize(m = mean(c(x, y, z)))
```

---
# List columns
If you apply rowwise operation with a list column, you don't have to loop

```{r }
df <- tibble(x = list(1, 2:3, 4:6))
```

.pull-left[
```{r }
df %>% 
  mutate(
    l = map_int(x, length)
  )
```
]

.pull-right[
```{r }
df %>% 
  rowwise() %>% 
  mutate(l = length(x))
```
]

---
# Creating list columns
You can use the `dplyr::nest_by()` function to create a list column for each group, *and* convert it to a rowwise data frame.

--
```{r }
d %>% 
  nest_by(content)
```

---
# Challenge
Given what we just learned, can you fit a model of the form `Theta ~ primary` to each content area (i.e., *not* using **{purrr}**)?

`r countdown::countdown(2)`

--
Wrap it in `list()` (should suggest this in the error reporting if you don't)

```{r }
d %>% 
  nest_by(content) %>% 
  mutate(m1 = list(lm(Theta ~ primary, data = data)))
```

---
# Challenge 2
Can you extend it further and extract the coefficients with `coef`? What about creating a new column that has the intercept values?

`r countdown::countdown(2)`

--
```{r }
d %>% 
  nest_by(content) %>% 
  mutate(m1 = list(lm(Theta ~ primary, data = data)),
         coefs = list(coef(m1)))
```

---
# Return atomic vectors

```{r }
d %>% 
  nest_by(content) %>% 
  mutate(m1 = list(lm(Theta ~ primary, data = data)),
         intercept = coef(m1)[1])
```

---
# Fit all models

The below gets us the same results we got before

```{r }
mods2 <- d %>%
  nest_by(content) %>% 
  mutate(
    m1 = list(lm(Theta ~ primary, data = data)),
    m2 = list(lm(Theta ~ primary + secondary, data = data)),
    m3 = list(lm(Theta ~ primary * secondary, data = data))
)
mods2
```

---
# Look at all $R^2$
### It's a normal data frame!

```{r gather-models}
mods %>%
  pivot_longer(
    m1:m3,
    names_to = "model",
    values_to = "output"
)
```

---
# Extract all $R^2$

*Note - might want to write a function here again*

```{r extract-r2}
r2 <- mods %>%
  pivot_longer(
    m1:m3,
    names_to = "model",
    values_to = "output"
  ) %>% 
  mutate(r2 = map_dbl(output, ~summary(.x)$r.squared))
r2
```

---
# Plot

```{r model-plot, fig.height = 6.5}
ggplot(r2, aes(model, r2)) +
	geom_col(aes(fill = model)) +
	facet_wrap(~content) +
	guides(fill = "none") +
	scale_fill_brewer(palette = "Set2")
```

---
# Unnesting

* Sometimes you just want to `unnest`

--

* Imagine we want to plot the coefficients by model... how?

--
* `broom::tidy()` => `tidyr::unnest()`



---
# Tidy

```{r}
mods %>%
	pivot_longer(
	  m1:m3,
	  names_to = "model", 
	  values_to = "output"
  ) %>%
	mutate(tidied = map(output, broom::tidy))
```

---
# Equivalently

```{r}
mods %>%
	pivot_longer(
	  m1:m3,
	  names_to = "model", 
	  values_to = "output"
  ) %>%
  rowwise() %>% 
  mutate(tidied = list(broom::tidy(output)))
```

---
# Select and unnest

```{r tidied}
tidied <- mods %>%
	gather(model, output, m1:m3) %>%
	mutate(tidied = map(output, broom::tidy)) %>%
	select(content, model, tidied) %>%
	unnest(tidied)
tidied
```

---
# Plot
### Lets look how the primary coefficients change

```{r coef-plot-echo, eval = FALSE}
to_plot <- names(coef(mods$m1[[1]]))

tidied %>%
  filter(term %in% to_plot) %>%
  ggplot(aes(estimate, term, color = model)) +
  geom_point() +
  scale_color_brewer(palette = "Set2") +
  facet_wrap(~content)
```

---
```{r coef-plot-eval, echo = FALSE, fig.height = 11}
to_plot <- names(coef(mods$m1[[1]]))

tidied %>%
	filter(term %in% to_plot) %>%
  ggplot(aes(estimate, term, color = model)) +
	geom_point() +
	scale_color_brewer(palette = "Set2") +
	facet_wrap(~content)
```

---
# Last bit
* We've kind of been running the wrong models this whole time

* We forgot about grade!

* No problem, just change the grouping factor

---
# By grade

```{r by_grade}
by_grade_content <- d %>%
  group_by(content, grade) %>% 
	nest()
by_grade_content
```

---
# Fit models

```{r by_grade_mods}
mods_grade <- by_grade_content %>%
  mutate(
    m1 = map(data, ~lm(Theta ~ primary, data = .x)),
    m2 = map(data, ~lm(Theta ~ primary + secondary, 
                       data = .x)),
    m3 = map(data, ~lm(Theta ~ primary * secondary, 
                       data = .x))
)
mods_grade
```

---
# Look at $R^2$

```{r by_grade_r2}
mods_grade %>%
	pivot_longer(
    m1:m3, 
    names_to = "model", 
    values_to = "output"
  ) %>% 
	mutate(r2 = map_dbl(output, ~summary(.x)$r.squared))
```

---
# Plot

```{r by_grade_r2_plot-echo, eval = FALSE}
mods_grade %>%
  pivot_longer(
    m1:m3, 
    names_to = "model", 
    values_to = "output"
  ) %>% 
  mutate(r2 = map_dbl(output, ~summary(.x)$r.squared)) %>%
  ggplot(aes(model, r2)) +
  geom_col(aes(fill = model)) +
  facet_grid(grade ~ content) +
  guides(fill = "none") +
  scale_fill_brewer(palette = "Set2")
```

---
```{r by_grade_r2_plot-eval, echo = FALSE, fig.height = 11.5}
mods_grade %>%
  pivot_longer(m1:m3, names_to = "model", values_to = "output") %>% 
  mutate(r2 = map_dbl(output, ~summary(.x)$r.squared)) %>%
  ggplot(aes(model, r2)) +
  geom_col(aes(fill = model)) +
  facet_grid(grade ~ content) +
  guides(fill = "none") +
  scale_fill_brewer(palette = "Set2")
```

---
# Summary
* List columns are really powerful and really flexible

* Also help you stay organized

* You can approach the problem either with **{purrr}** or `dplyr::rowwise()`. 

  + **Important**: If you use `rowwise()`, remember to `ungroup()` when you want it to go back to being a normal data frame
  
  + I'm asking you to learn both - the row-wise approach might be a bit easier but is a little less general (only works with data frames)


---
class: inverse-blue middle

# Questions?

---
class: inverse-green middle
# In-class Midterm
### Next time: Parallel iterations


