---
title: "Looping Variants"
author: "Daniel Anderson "
date: "Week 5, Class 2"
output:
  xaringan::moon_reader:
    css: ["default", "new.css"]
    lib_dir: libs
    nature:
      navigation:
        scroll: false
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
    includes:
      in_header: "load-feather.html"
---

```{r include = FALSE, results = "asis"}
source(here::here("static", "slides", "slide-setup.R"))
xaringanExtra::use_clipboard()
knitr::opts_chunk$set(fig.width = 13, 
                      message = FALSE, 
                      warning = FALSE)

library(tidyverse)
theme_set(theme_minimal(20))
update_geom_defaults("point", list(size = 3))
```

`r setup("w5p2")`

---
# Agenda
* `walk()` and friends

* `modify()`

* `safely()`

* `reduce()`

---
# Learning Objectives

* Know when to apply `walk` instead of `map`, and why it may be useful

* Understand the parallels and differences between `map` and `modify`

* Diagnose errors with `safely` and understand other situations where it may
be helpful

* Collapsing/reducing lists with `purrr::reduce()` or `base::Reduce()`



---
# Setup
Let's go back to our plotting example from last class.

--

First we'll load our libraries

```{r }
library(tidyverse)
library(fivethirtyeight)
library(glue)
library(english)
```

---
# Prep the data

```{r }
pulitzer <- pulitzer %>%
  select(newspaper, starts_with("num")) %>%
  pivot_longer(
    -newspaper,
    names_to = "year_range",
    values_to = "n",
    names_prefix = "num_finals"
  ) %>% 
  mutate(year_range = str_replace_all(year_range, "_", "-")) %>%
  filter(year_range != "1990-2014") %>% 
	group_by(newspaper) %>%
	mutate(
	  tot = sum(n),
	  label = glue(
	    "{str_to_title(as.english(tot))} Total Pulitzer Awards"
    )
)
```


---
# Produce plots
```{r }
final_plots <- pulitzer %>%
	group_by(newspaper, label) %>%
	nest() %>%
	mutate(plots = pmap(list(newspaper, label, data), ~{
    ggplot(..3, aes(n, year_range)) + 
      geom_col(aes(fill = n)) +
      scale_fill_distiller(type = "seq", 
                           limits = c(0, max(pulitzer$n)),
                           palette = "BuPu",
                           direction = 1) +
	    scale_x_continuous(limits = c(0, max(pulitzer$n)), 
	                       expand = c(0, 0)) +
	    guides(fill = "none") +
	    labs(title = glue("Pulitzer Prize winners: {..1}"),
	         x = "Total number of winners",
	         y = "",
	         caption = ..2)
	  })
)
```

---
# Saving
* We saw last time that we could use `nest_by()`
  + Required a bit of awkwardness with adding the paths to the data frame
  
--
* Instead, we'll do it again but with the `walk()` family


--
Why `walk()` for saving instead of `map()`?

> Walk is an alternative to map that you use when you want to call a function for its side effects, rather than for its return value. You typically do this because you want to render output to the screen or save files to disk - the important thing is the action, not the return value.

\-[r4ds](https://r4ds.had.co.nz/iteration.html#walk)

---
class: inverse-red
# More practical
If you use `walk()`, nothing will get printed to the screent. This is particularly helpful for RMarkdown files.

---
# Example
Please do the following
* Create a new RMarkdown document
* Paste the code you have for creating the plots in a code chunk there (along with the library loading, data cleaning, etc.)

`r countdown::countdown(1)`

---
# Create a directory
```{r create_dir}
fs::dir_create(here::here("plots", "pulitzers"))
```

--
### Create file paths
```{r create_file-paths}
newspapers <- str_replace_all(tolower(final_plots$newspaper), " ", "-")
paths <- here::here("plots", "pulitzers", glue("{newspapers}.png"))
```

---
# Challenge
* Use a `map()` family function to loop through `paths` and `final_plots$plots` to save all plots. 

* Render (knit) your file. What do you notice?

`r countdown::countdown(3)`

---
# `walk()`
Just like `map()`, we have parallel variants of `walk()`, including, `walk2()`, and `pwalk()`

These work just like `map()` but don't print to the screen

Try replacing your prior code with a `walk()` version. 

How does the rendered output change?

`r countdown::countdown(2)`


---
# Save plots

```{r walk2, eval = FALSE}
walk2(paths, final_plots$plots, ggsave,
      width = 9.5, 
      height = 6.5,
      dpi = 500)
```

---
class: inverse-red middle
# modify

---
class: middle

> Unlike `map()` and its variants which always return a fixed object type (list for `map()`, integer vector for `map_int()`, etc), the `modify()` family always returns the same type as the input object.

---
# `map` vs `modify`
### map

```{r map-mtcars}
map(mtcars, ~as.numeric(scale(.x)))
```

---
### modify

```{r modify-mtcars}
modify(mtcars, ~as.numeric(scale(.x)))
```


---

```{r map-paste}
modify2(LETTERS[1:3], letters[1:3], paste0)
map2(LETTERS[1:3], letters[1:3], paste0)
```

---
class: inverse-red middle
# safely

---
## Iterating when errors are possible

Sometimes a loop will work for most cases, but return an error on a few


--
Often, you want to return the output you can


--
Alternatively, you might want to diagnose *where* the error is occurring


--
`purrr::safely`


---
# Example
```{r by_cyl}
by_cyl <- mpg %>% 
  group_by(cyl) %>% 
  nest() 
by_cyl
```


--

```{r lm-fail, error = TRUE}
by_cyl %>% 
  mutate(mod = map(data, ~lm(hwy ~ displ + drv, data = .x)))
```

---
# Safe return

* First, define safe function - note that this will work for .ital[.b[any]] function
```{r safe-lm}
safe_lm <- safely(lm)
```


--

* Next, loop the safe function, instead of the standard function


--

```{r loop-safe-lm}
safe_models <- by_cyl %>% 
  mutate(safe_mod = map(data, ~safe_lm(hwy ~ displ + drv, data = .x)))
safe_models
```

---
# What's returned?

```{r safe-return}
safe_models$safe_mod[[1]]
safe_models$safe_mod[[4]]
```


---
# Inspecting
I often use `safely()` to help me de-bug. Why is it failing *there*.


--
First - create a new variable to filter for results with errors

--
```{r }
safe_models %>% 
  mutate(error = map_lgl(safe_mod, ~!is.null(.x$error)))
```

---
# Inspecting the data
```{r }
safe_models %>% 
  mutate(error = map_lgl(safe_mod, ~!is.null(.x$error))) %>% 
  filter(error) %>% 
  select(cyl, data) %>% 
  unnest(data)
```

The `displ` and `drv` variables ar constant, so no relation can be estimated.


---
# Pull results that worked

```{r results}
safe_models %>% 
  mutate(results = map(safe_mod, "result"))
```

--
Now we can `broom::tidy()` or whatevs

---

Notice that there is no `cyl == 5`.

```{r }
safe_models %>% 
  mutate(results = map(safe_mod, "result"),
         tidied = map(results, broom::tidy)) %>% 
  select(cyl, tidied) %>% 
  unnest(tidied)
```

---
# When else might we use this?

--

Any sort of web scraping - pages change and URLs don't always work

---
# Example

```{r error = TRUE}
library(rvest)
links <- list(
  "https://en.wikipedia.org/wiki/FC_Barcelona",
  "https://nosuchpage",
  "https://en.wikipedia.org/wiki/Rome"
)
pages <- map(links, ~{
  Sys.sleep(0.1)
  read_html(.x)
})
```

---
# The problem
I can't connect to https://nosuchpage because it doesn't exist


--
.center[.blue[.realbig[BUT]]]


--
That also means I can't get *any* of my links because *one* page errored (imagine it was 1 in 1,000 instead of 1 in 3)


--
## `safely()` to the rescue

---
# Safe version

```{r }
safe_read_html <- safely(read_html)
pages <- map(links, ~{
  Sys.sleep(0.1)
  safe_read_html(.x)
})
str(pages)
```

---
# Non-results
In a real example, we'd probably want to double-check the pages where we got no results

--
```{r }
errors <- map_lgl(pages, ~!is.null(.x$error))
links[errors]
```


---
class: inverse-red
# reduce

---
# Reducing a list
The `map()` family of functions will always return a vector the same length as the input

--
`reduce()` will collapse or reduce the list to a single element

---
# Example

```{r }
l <- list(
  c(1, 3),
  c(1, 5, 7, 9),
  3,
  c(4, 8, 12, 2)
)

reduce(l, sum)
```

---
# What's going on?
The code `reduce(l, sum)` is the same as

```{r }
sum(l[[4]], sum(l[[3]], sum(l[[1]], l[[2]])))
```

Or slidghlty differently
```{r }
first_sum <- sum(l[[1]], l[[2]])
second_sum <- sum(first_sum, l[[3]])
final_sum <- sum(second_sum, l[[4]])
final_sum
```

---
# Why might you use this?
What if you had a list of data frames like this

```{r }
l_df <- list(
  tibble(id = 1:3, score = rnorm(3)),
  tibble(id = 1:5, treatment = rbinom(5, 1, .5)),
  tibble(id = c(1, 3, 5, 7), other_thing = rnorm(4))
)
```

We can join these all together with a single loop - we want the output to be of length 1!

---
```{r }
reduce(l_df, full_join)
```

--
Note - you have to be careful on directionality

.pull-left[
```{r }
reduce(l_df, left_join)
```
]

.pull-left[
```{r }
reduce(l_df, right_join)
```
]

---
# More common
You probably just want to `bind_rows()`

```{r }
l_df2 <- list(
  tibble(id = 1:3, scid = 1, score = rnorm(3)),
  tibble(id = 1:5, scid = 2, score = rnorm(5)),
  tibble(id = c(1, 3, 5, 7), scid = 3, score = rnorm(4))
)
reduce(l_df2, bind_rows)
```

---
# Wrap up
* Lots more to `{purrr}` but we've covered a lot
  
* Functional programming can *really* help your efficiency, and even if it slows you down initially, I'd recommend always striving toward it, because it will ultimately be a huge help.


.center[

### Questions?

]

--
If we have any time left - let's work on the homework