---
title: "Shiny: Part 2"
author: "Daniel Anderson "
date: "Week 8, Class 2"
output:
  xaringan::moon_reader:
    css: ["default", "uo", "uo-fonts", "custom.css"]
    lib_dir: libs
    nature:
      highlightStyle: atelier-dune-light
      highlightLines: true
      countIncrementalSlides: false
      beforeInit: "https://platform.twitter.com/widgets.js"
---

```{r setup, include = FALSE}
library(tidyverse)
theme_set(theme_minimal(20))
update_geom_defaults("point", list(size = 3))
knitr::opts_chunk$set(fig.width = 13, 
                      message = FALSE, 
                      warning = FALSE)
```

# Agenda
* Review Lab 4

* Review shiny basics

* Discuss some layout options

* Introduce shiny dashboard

* Some extensions worth exploring later

---
# Learning objectives
* Solidify your understanding of the UI/Server relation

* Be able create tabsets and a navbar

* Be able to create basic shiny dashboards

---
class: inverse center middle
# Review lab


---
# Basic shiny

Create a basic shiny template, as we did before.

`r countdown::countdown(minutes = 3, 
                        left = 0, 
                        right = 0,
                        bottom = 0.5,
                        color_running_background = "#FEE11A",
                        color_running_text = "#000")`


---
# UI 

* The `ui` defines the look and feel of the app - the user interface

* Use it to define where output "lives"

* It also defines the inputs for the `server`, where functions are actually
evaluated.

* In the template/default case, a `sliderInput` has been defined, which we're
calling `"bins"`. It will take on values from 1 to 50, and will start at 30.

* Access the specific value the user selects within the `server`, through `input$bins`.


---
# Server

* The `server` function takes the `input` from the UI and produces `output` with normal R code.

* In this case, we're creating one output object, called `distPlot`. The result is then called through the `ui` on line 30


---
# Change the input
Let's switch from a slider to a drop down menu. 


--
### How?

Even if you don't know the specific code, what would we change?


--
`sliderInput` will become `selectInput`. The arguments will also be slightly
  different


--
### Try!

`r countdown::countdown(minutes = 2, 
                        left = 0, 
                        right = 0,
                        bottom = 1,
                        color_running_background = "#FEE11A",
                        color_running_text = "#000")`

---
class: inverse center middle

# layouts

---
# Tabs

* Let's say we wanted a tabset with different things.


--
* First, we need at least two things!


--
* Let's create a table that has the lower/upper bound of the bin, and the counts
within that range.

---
# Table creation
.footnote[Because this is base syntax, I'll give you the basics for the table, you focus
on the shiny part]

```{r tbl, eval = FALSE}
input_bins <- 30 # placeholder for whatever the input is

x <- faithful[, 2] # data you're working with

bins <- seq(min(x), 
            max(x), 
            length.out = input_bins + 1)

h <- hist(x, breaks = bins, plot = FALSE)

tibble(lower = lag(h$breaks),
       upper = h$breaks) %>%
  drop_na(lower) %>%
  mutate(counts = h$counts) %>%
  mutate_if(is.numeric, round, 2)
```
--
* You take it from here! Add a table below the plot with `DT::datatable` or `reactable::reactable`

`r countdown::countdown(minutes = 8, 
                        left = 1, 
                        right = 0,
                        bottom = 0,
                        top = 1,
                        color_running_background = "#FEE11A",
                        color_running_text = "#000")`

---
class: inverse center middle

# [demo]


---
# Move it to a tabset
* Just create a `tabsetPanel` within the `mainPanel`, then put the output for each tab within `tabPanel`.

`r countdown::countdown(minutes = 5, 
                        left = 0, 
                        right = 0,
                        bottom = 0.5,
                        color_running_background = "#FEE11A",
                        color_running_text = "#000")`




---
# Different pages
### Add a navbar

* Instead of using a tabset with `tabsetPanel`, you might want to have a navbar at the top of the page, which you can create with `navbarPage`.


--
* Can be a bit more complicated - each `tabset` needs to include everything, including the `sidebarPanel` (if present), could include tabsets, `mainPanel`, etc.


--
* Essentially each tab from the `navbar` becomes an entirely new page.

---
# Important things with navbar

* Move the title of your application to the first argument of `navbar` .g[otherwise you'll end up with weird renderings]


--
* Use separate input IDs even if you're repeating the same thing (otherwise it won't be responsive)


---
class: inverse center middle


# [demo]


---
# More on the navbar
* Can really help with organization/flexibility (you could even have tabs
within a page)

* Refactoring can help organization A LOT
  + Pull pieces out to try to make code more readable/less buggy.



---
class: inverse center middle
# {shinydashboard}
[Getting started](https://rstudio.github.io/shinydashboard/get_started.html)

---
### First dashboard - ui
```{r dashboard1-ui, eval = FALSE}
library(shiny)
library(shinydashboard)

ui <- dashboardPage(
  dashboardHeader(title = "Basic dashboard"),
  dashboardSidebar(),
  dashboardBody(
    # Boxes need to be put in a row (or column)
    fluidRow(
      box(plotOutput("plot1", height = 250)),

      box(
        title = "Controls",
        sliderInput("slider", "Number of observations:", 1, 100, 50)
      )
    )
  )
)
```

---
# First dashboard - server

```{r dashboard1-server, eval = FALSE}
server <- function(input, output) {
  set.seed(122)
  histdata <- rnorm(500)

  output$plot1 <- renderPlot({
    data <- histdata[seq_len(input$slider)]
    hist(data)
  })
}
```

--
### Run it

```{r, run-dashboard, eval = FALSE}
shinyApp(ui, server)
```

[demo]

---
# Main differences
* You now have `dashboardSidebar` and `dashboardBody`

* You also now have `fluidRow` and `box` arguments to arrange things in the main
body


---
# Sidebar
* Probably the defining characteristic of the dashboard

  + Define a `sidebarMenu` with `menuItem`s


--
### Example

```{r sidebar, eval = FALSE}
sidebarMenu(
  menuItem("Histogram", tabName = "histo", icon = icon("chart-bar")),
  menuItem("Bin Counts", tabName = "bins", icon = icon("table"))
    )
```

You can also do things like put the slider in the `sidebarMenu`

[demo]

---
# Referencing menu items

* If you define `menuItem`s, you'll have to give them a `tabName` (see previous
slide).

* In the `dashboardBody`, create a `tabItems` with specific `tabItem` pieces.
  This should be how you control/refer to the `menuItem`.


[demo]



---
# Extension

* There's lots of extensions for shiny, and quite a few (but not as many) for shinydashboard

* Consider themeing shiny apps with [{shinythemes}](https://rstudio.github.io/shinythemes/) and dashboards with [{dashboardthemes}](https://github.com/nik01010/dashboardthemes)

* Consider themeing figures to match your shiny theme with [{thematic}](https://rstudio.github.io/thematic/)

---
# Conclusions
* Shiny is super customizable - almost limitless (see more examples [here](https://shiny.rstudio.com/gallery/))

* Great for building interactive plots, but you can use it for all sorts of other things too (including text and tables)

* Really helpful and fun way to build data tools for practitioners

* Takes some practice, but basically allows you to write normal R code, and get interactive websites